---
title: "The Netherlands as a Part-Time Economy"
author: "Hidde Hendrickx (2831619), Andrew Messih (2785753)"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 7, fig.height = 4)

options(vroom.show_progress = FALSE, readr.show_progress = FALSE)

# install.packages(c("tidyverse", "eurostat", "sf", "giscoR"))
library(tidyverse)
library(eurostat)
library(sf)
```

# 1. Identification of the Social Problem

## 1.1 Describe the Social Problem

The Netherlands has the highest rate of part-time work in Europe. Roughly 39% of Dutch
workers have a part-time job (OECD, 2024). At the same time, the country has one of the
highest overall employment rates in the EU. This means the Dutch labour market is not about
a group of people working long hours, but rather a large group of people working fewer
hours.

Part-time work in the Netherlands is strongly connected to gender. About three out of four
working women have a part-time job, compared to roughly one out of five working men
(Portegijs & van den Brakel, 2016). This causes debates about part-time work in the
Netherlands to include debates about women, labour market participation and economic
independence.

The Netherlands is very different from other European countries. In 2024, Dutch workers
between the age of 20 and 64 worked an average of 32.1 hours per week, against an EU average
of 36.0 hours per week (Eurostat, 2025). With current labour shortages, this is relevant
for policymakers; if a large share of the workforce works part-time, it might be possible to
increase hours worked.

## 1.2 Research Question and Sub-Questions

**Main question:** How has part-time employment in the Netherlands developed over time,
and how does it differ between men and women?

Sub-questions:

1. How has the share of part-time workers in the Netherlands changed over time?
   *(temporal variation)*
2. What are the differences in part-time employment between men and women?
   *(sub-population variation)*
3. How does the Netherlands compare to other European countries? *(spatial variation)*
4. Did part-time employment change around the COVID period? *(event analysis)*

# 2. Data Sourcing

## 2.1 Description of the Datasets

We use data from Eurostat, the statistical office of the European Union. The main dataset
is part-time employment as a percentage of total employment, by sex and age (Eurostat code
lfsa_eppga). It contains, per country and per year, the share of employed people who work
part-time. We download it directly into R with the eurostat package, giving anyone who runs
this file the ability to reproduce the analysis.

```{r load-data, results = "hide"}
pt_raw <- get_eurostat("lfsa_eppga", time_format = "num")
pt_raw <- rename(pt_raw, year = TIME_PERIOD)
```

## 2.2 Data Sources and Limitations

The data come from the EU Labour Force Survey, a large household survey done in all EU
member states. There are some limitations to our analysis. First, the data is from a survey,
so the numbers are estimates with a margin of error. Second, the definition of part-time
work varies across countries; most countries rely on what respondents say themselves, but
the Netherlands uses a 35-hour limit, which results in the Dutch figures being relatively
high compared to sources that use a 30-hour rule such as the OECD. Third, in 2021, a new EU
regulation was introduced which changed the survey methodology, causing a break in the time
series (we come back to this in the event analysis).

To combine more than one source, we add a second table that groups each country into a
European region. This allows us to compare the Netherlands to whole regions instead of just
countries.

```{r region-table}
regions <- tribble(
  ~geo, ~region,
  "NL", "Western Europe",  "BE", "Western Europe",  "DE", "Western Europe",
  "FR", "Western Europe",  "LU", "Western Europe",  "AT", "Western Europe",
  "DK", "Northern Europe", "SE", "Northern Europe", "FI", "Northern Europe",
  "IE", "Northern Europe", "ES", "Southern Europe", "IT", "Southern Europe",
  "PT", "Southern Europe", "EL", "Southern Europe", "PL", "Eastern Europe",
  "CZ", "Eastern Europe",  "HU", "Eastern Europe",  "RO", "Eastern Europe",
  "SK", "Eastern Europe"
)
```

## 2.3 Description of Variables

| Variable   | Meaning                                          |
|------------|--------------------------------------------------|
| geo        | Country (Eurostat 2-letter code, e.g. NL)        |
| sex        | T = total, M = male, F = female                  |
| age        | Age group (we keep 15--64)                       |
| year       | Year of observation                              |
| pt_rate    | Part-time workers as % of total employment       |
| region     | European region (added by us)                    |

# 3. Quantifying Part-Time Employment

## 3.1 Data Cleaning

We clean the raw data in a few clear steps:

1. We only keep the working-age group (15--64) and the percentage unit.
2. We drop rows where the part-time rate is missing, because that is our outcome.
3. We keep only the columns we need and give them readable names.
4. We merge in the region table from Section 2.2.

```{r clean-data}
pt_clean <- pt_raw %>%
  filter(age == "Y15-64", unit == "PC") %>%
  filter(!is.na(values)) %>%
  select(geo, sex, year, pt_rate = values) %>%
  left_join(regions, by = "geo")
```

## 3.2 Creation of New Variables

We made two new variables that we will use in the plots below.

**Gender gap.** For every country and year we compute the difference between the female and
male part-time rate. If a number is positive, it means women work part-time more often than
men. We also compute a ratio (female rate divided by male rate). For example, a ratio of 3
means women work part-time three times as often as men. We use the gap in Section 3.4.

**Change around 2021.** For each country we compare the average part-time rate just before
(2018--2020) and just after (2021--2023) the break with each other. This shows us how big
the change was around 2021, which we use in Section 3.6.

```{r new-variables}
gap <- pt_clean %>%
  filter(sex %in% c("M", "F")) %>%
  mutate(sex = recode(sex, M = "Male", F = "Female")) %>%
  select(geo, year, sex, pt_rate) %>%
  pivot_wider(names_from = sex, values_from = pt_rate) %>%
  mutate(gender_gap = Female - Male,
         gender_ratio = Female / Male)

change <- pt_clean %>%
  filter(sex == "T", year %in% 2018:2024) %>%
  mutate(period = if_else(year <= 2020, "before", "after")) %>%
  group_by(geo, period) %>%
  summarise(mean_rate = mean(pt_rate), .groups = "drop") %>%
  pivot_wider(names_from = period, values_from = mean_rate) %>%
  mutate(change_2021 = after - before)
```

## 3.3 Temporal Variation

```{r plot-temporal}
pt_clean %>%
  filter(geo == "NL", sex == "T") %>%
  ggplot(aes(year, pt_rate)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "steelblue") +
  labs(title = "Part-time employment in the Netherlands over time",
       x = "Year", y = "Part-time workers (% of total employment)") +
  theme_minimal()
```

The chart shows that part-time employment in the Netherlands rose fairly constant from about
21% of total employment in 1983 to roughly 50% by the early 2010s. From then until 2020 the
rate remained quite stable, peaking at about 51% in 2020. In 2021 it drops sharply to around
43% and almost stays the same in the years after. This break should be read with care; the
EU Labour Force Survey changed its methodology in 2021, so the 2021 fall could indicate a
measurement effect rather than a real decline in part-time work.

## 3.4 Gender Differences in Part-Time Employment

```{r plot-gender}
pt_clean %>%
  filter(geo == "NL", sex %in% c("M", "F")) %>%
  mutate(sex = recode(sex, M = "Male", F = "Female")) %>%
  ggplot(aes(year, pt_rate, colour = sex)) +
  geom_line(linewidth = 1) +
  geom_point() +
  labs(title = "Part-time employment by sex, Netherlands",
       x = "Year", y = "Part-time workers (% of total employment)", colour = "Sex") +
  theme_minimal()
```

```{r plot-gendergap}
gap %>%
  filter(geo == "NL") %>%
  ggplot(aes(year, gender_gap)) +
  geom_line(colour = "darkorchid", linewidth = 1) +
  geom_point(colour = "darkorchid") +
  labs(title = "Gender gap in part-time employment, Netherlands",
       x = "Year", y = "Gender gap (female - male, % points)") +
  theme_minimal()
```

Across the whole period women work part-time far more often than men. The female rate climbs
from about 50% in 1983 to a peak of roughly 76% in the mid-2010s, while the male rate rises
from under 10% to about 28% by 2020. The gap is therefore large and persistent: in 2024
around 63% of women work part-time against about 24% of men, a difference of roughly 40
percentage points (women work part-time about 2.6 times as often as men). Both lines fall in
2021, in line with the methodological break discussed below, but the gender gap stays wide
throughout. This pattern reflects the social problem: part-time work in the Netherlands is
mostly done by women, which relates directly to discussions about women's participation in
the labour-market and economic independence.

## 3.5 Spatial Variation Across Europe

```{r plot-map}
latest <- pt_clean %>%
  filter(sex == "T", year == max(year))

europe <- get_eurostat_geospatial(resolution = "20", nuts_level = "0", year = "2021")

map_df <- left_join(europe, latest, by = "geo")

ggplot(map_df) +
  geom_sf(aes(fill = pt_rate)) +
  coord_sf(xlim = c(-12, 35), ylim = c(34, 71)) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(title = "Part-time employment across Europe (latest year)",
       fill = "Part-time %", caption = "Source: Eurostat (lfsa_eppga)") +
  theme_minimal()
```

The map shows that the Netherlands has the highest part-time employment rate across Europe.
Other countries in Northwestern Europe also have a high part-time employment rate, while the
countries in Southeastern Europe have a low part-time employment rate.

## 3.6 Event Analysis

Instead of just looking at the Netherlands, we compare it with a few neighbouring countries
around 2020--2021, and we use our change variable to measure the size of the jump in each
country.

```{r plot-event}
pt_clean %>%
  filter(geo %in% c("NL", "DE", "FR", "BE"), sex == "T", year >= 2010) %>%
  mutate(country = recode(geo, NL = "Netherlands", DE = "Germany",
                          FR = "France", BE = "Belgium")) %>%
  ggplot(aes(year, pt_rate, colour = country)) +
  geom_line(linewidth = 1) +
  geom_point() +
  geom_vline(xintercept = 2021, linetype = "dashed") +
  labs(title = "Part-time employment around 2021, selected countries",
       x = "Year", y = "Part-time workers (% of total employment)", colour = "Country",
       caption = "Dashed line: 2021 EU-LFS methodological break (also the COVID period)") +
  theme_minimal()
```

```{r plot-change}
change %>%
  filter(geo %in% c("NL", "DE", "FR", "BE")) %>%
  mutate(country = recode(geo, NL = "Netherlands", DE = "Germany",
                          FR = "France", BE = "Belgium")) %>%
  ggplot(aes(country, change_2021, fill = country)) +
  geom_col() +
  labs(title = "Change in part-time rate around 2021 (2021-2023 vs 2018-2020)",
       x = NULL, y = "Change (% points)") +
  theme_minimal() +
  theme(legend.position = "none")
```

Around 2021 the Dutch line drops sharply, from about 50% to roughly 43%, while Germany,
France and Belgium barely move. Because COVID-19 affected all four countries, the fact that
only the Netherlands shows a jump makes COVID an unlikely explanation; a real pandemic
effect would be expected to appear in the neighbouring countries too. The drop lines up
exactly with the 2021 EU-LFS methodological change, which reclassified many Dutch workers
because of the country's 35-hour part-time definition. We interpret the 2021 movement as an
association with a measurement change, not as evidence that COVID caused a real fall in
part-time work.

# 4. Discussion

## 4.1 Main Findings

Part-time employment in the Netherlands rose from about 21% of total employment in 1983 to
roughly 50% by the early 2010s and then stayed broadly stable until 2020. It is strongly
gendered; in 2024 around 63% of women work part-time against about 24% of men. Across Europe
the Netherlands has the highest rate, with Northwestern Europe also high and Southeastern
Europe low. The sharp drop in 2021 does not appear in Germany, France or Belgium, so it's
most likely that it reflects the 2021 EU-LFS methodological change rather than COVID-19.

## 4.2 Limitations

The data are survey estimates, the definition of part-time work differs between countries
(the Netherlands uses a 35-hour limit), and the 2021 methodological break makes it hard to
compare years before and after 2021. Our event analysis therefore shows correlation, not
causation. The region table is a simplification of European geography.

## 4.3 Implications

From a policy perspective, a large and overwhelmingly female part-time workforce means there
is room to raise hours worked amid labour shortages. However, this is shaped by childcare,
taxation and norms, and our findings show association, not causation.

# 5. Reproducibility

## 5.1 GitHub Repository Link

All code for this report is available in our public GitHub repository:

https://github.com/HIddehendrickx/Programming-for-Economist-

## 5.2 How to Reproduce the Analysis

1. Open this .Rmd file in RStudio.
2. Install the required packages (only the first time):
   `install.packages(c("tidyverse", "eurostat", "sf", "giscoR"))`
3. Press Knit. All data is downloaded automatically from Eurostat, so no files need to be
   added by hand.

# 6. References

- Eurostat. (2025). *EU labour market — working hours.* https://ec.europa.eu/eurostat/web/products-eurostat-news/w/ddn-20250514-1
- Eurostat. *Part-time employment as percentage of the total employment, by sex and age (lfsa_eppga).* EU Labour Force Survey.
- OECD. (2024). *Part-time employment rate.* https://www.oecd.org/en/data/indicators/part-time-employment-rate.html
- Portegijs, W., & van den Brakel, M. (2016). *Emancipatiemonitor 2016.* Centraal Bureau voor de Statistiek (CBS). https://www.cbs.nl
